\bf Is this normal? A new projection pursuit index to assess a sample against a multivariate null distribution

### 2024-07-05 paper.qmd
Comparison with Shapley-Outlier result added.

```{r}
#| eval: false
#| echo: true
library(tourr)
library(mulgar)
set.seed(929)
vc_null <- matrix(rep(0.5, 5*5), ncol=5)
diag(vc_null) <- 1
m_null <- rep(0, 5)
vc_samp <- matrix(rep(0, 5*5), ncol=5)
diag(vc_samp) <- 1
vc_samp[4,5] <- -0.47
vc_samp[5,4] <- -0.56
vc_samp <- vc_samp*0.1
m_samp <- c(0, 0, 0, 1.9, 2.3)
samp <- as.data.frame(rmvn(6, 
                           mn = m_samp, 
                           vc = vc_samp))
animate_xy(samp, guided_anomaly_tour(anomaly_index(),
  ellipse=vc_null), ellipse=vc_null, 
  axes = "bottomleft", half_range=5, center=FALSE)
```

@@

Here we will consider points to be outlying if they are more than $5 \sigma$ away from that mean value. With $p=16$ this corresponds to a Mahalanobis distance of $60$ or larger. This will identify $20$ out of the $68$ observations as outlying. Since these are outlying in different combinations of variables, as found in @filzmoser2018, we will further separate the outlying points into clusters based on  similarity in direction. The similarity is computed by first normalizing observations to have length $1$ and then apply k-means clustering with Euclidean distance. The cluster validation statistics computed using @fpc suggest that $k=4$ or $k=5$ is reasonable, so for simplicity we will work with $k=4$.

The new anomaly index is first applied to the full dataset, such that the final projection will be affected by averaging distance of points in many directions. It provides a global picture showing where the outlying points differ from the normal distribution.

@@

# extract final view, draw with new axis display that
# only labels longer arrows, draw without red X

# cluster 4 only has two points, they are more in
# the center when optimizing for all outliers
# next step: check optimization for only cluster 4
# also cluster 3 might be of interest, it appears
# somewhat spread out in the final view, but it is
# also the largest cluster

@@

XXX SOMETHING ON WHAT THIS MEANS RELATIVE TO THE ORIGINAL EXAMPLE

>>>>

```{r}
#| eval: false
#| echo: false
set.seed(929)
vc_null <- matrix(rep(0.5, 5*5), ncol=5)
diag(vc_null) <- 1
m_null <- rep(0, 5)
vc_samp <- matrix(rep(0, 5*5), ncol=5)
diag(vc_samp) <- 1
vc_samp[4,5] <- -0.47
vc_samp[5,4] <- -0.56
vc_samp <- vc_samp*0.1
m_samp <- c(0, 0, 0, 1.9, 2.3)
samp <- as.data.frame(rmvn(6, 
                           mn = m_samp, 
                           vc = vc_samp))
```

```{r}
#| eval: false
#| echo: true
library(tourr)
animate_xy(samp, guided_anomaly_tour(anomaly_index(),
  ellipse=vc_null), ellipse=vc_null, 
  axes = "bottomleft", half_range=5, center=FALSE)
```

@@

Here we will consider points to be outlying if they are more than $5 \sigma$ away from that mean value. With $p=16$ this corresponds to a Mahalanobis distance of $60$ or larger. This will identify $20$ out of the $68$ observations as outlying. Since these are outlying in different combinations of variables, as found in @filzmoser2018, we will further separate the outlying points into clusters based on  similarity in direction. The similarity is computed by first normalizing observations to have length $1$ and then apply k-means clustering with Euclidean distance. We use the Dunn index @dunn, computed using @fpc, to select the preferred number of clusters, $k=5$.

The new anomaly index is first applied to the full dataset, such that the final projection will be affected by averaging distance of points in many directions, see Fig. @fig-weather (a). It provides a global picture showing where the outlying points differ from the normal distribution. We can see that the contrast between maximum (atmx) and minimum (atmn) temperature is primarily used in the direction where points are most outlying. In the orthogonal direction the variables precipitation (prc) and number of days with maximum temperature above 25°C (nsm) can be used to separate the clusters 2, 3 and 5. With this solution the two points in cluster 4 cannot be resolved, they fall inside the cluster. To better understand what is different for those two years we separately run the anomaly index for this subset, see Fig. @fig-weather (b). From the result we see that the maximum temperature is not relevant for this cluster, but that they have low values for the minimum temperature. We also see that there is still some averaging, and the preferred solution is not placing either point on the outside of the projected cluster.

@@

@@

To compare our results to what is obtained by outlier detection methods, we reproduce one of the results from @filzmoser2018 in @fig-shapley. On top we can see the cluster assignments for all points that we have identified as outliers using the $p$-dimensional Mahalanobis distance.

For example we see that cluster 4 contains the years 1962 and 1996, and for these two years the result from @filzmoser2018 also indicates unusually low values for the minimum teperature, while there is disagreement between outlyingness in other variables. Note that compared to the years 1965 and 1966 they have unusually low minimum temperature while having average values for the number of days with maximum temparature above 25°C, and this contrast is also captured in the projection.

When looking at the overall picture (Fig. @fig-weather (a)), we can see that the linear projection is showing the difference between minimum and maximum temperature, which is where clusters are outlying in different directions. The direction of cluster 1 is where the average maximum temperature is unusually higher compared to the minimum temperature, while clusters 3 corresponds to years where the minimum temperature is unusually close to the maximum temperature. This combined information cannot be resolved with the cell detection algorithm, but we see that for example the most recent years (cluster 3) were tagged for having unusually high values for both the minimum and the maximum temperature.

```{r}
#| eval: false
#| echo: false
library(ShapleyOutlier)
Sigma_inv <- solve(MCD$cov)
# following the package vignette to calculate SCD
weather_SCD <- SCD(x = X, mu, Sigma, Sigma_inv, step_size = 0.1, min_deviation = 0.2)
weather_SCD_rescaled <- weather_SCD
weather_SCD_rescaled$x_original <- t(apply(weather_SCD_rescaled$x_original,1, function(x) x*attr(X, "scale") + attr(X, "center")))
weather_SCD_rescaled$x <- t(apply(weather_SCD_rescaled$x,1, function(x) x*attr(X, "scale") + attr(X, "center")))

# when drawing I want to add clustering information
labels <- data.frame(
  x = c(seq(1960, 2020, length.out = 7),
        as.numeric(names(X_km$cluster))),
  label = c(seq(1960, 2020, length.out = 7),
        paste0("\n",X_km$cluster))
)
cluster_df <- as_tibble(X_km$cluster, rownames = "year") %>%
  add_column(variable = "num_precp_01")
p <- plot(weather_SCD_rescaled, type = "cell", n_digits = 0, continuous_rowname = 10, rotate_x = FALSE) +
  geom_text(data = cluster_df,
            mapping = aes(x = as.numeric(year),
                               y = variable, label = value,
                          color = as.factor(value)),
            size = 2, fontface = "bold",
            inherit.aes = F) +
  scale_color_manual(values = c("#80146E", "#7262B2", "#31A0C2", "#5ECBBB", "#B9E5B9")) +
  theme(legend.position = "none",
        aspect.ratio = 0.5,
        text = element_text(size = 10)) +
  scale_x_continuous(breaks = seq(1960, 2020, 5),
                     labels = c(as.character(seq(60, 95, 5)),
                                "00", "05",
                                as.character(seq(10, 20, 5))))
ggsave("images/weather3.png", p, "png", bg = "white",
       width = 6, height = 3)
```

::: {#fig-shapley}

![](images/weather3.png){#fig-weather3 width=600 align="center"}

Reproduction of the results from @filzmoser2018, showing which cells have been identified as outlying by their Shapley Cell Detector algorithm. Red cells indicate values higher than expected, blue cells values lower than expected. For years that have been identified as outliers in our approach we show the cluster assignment at the top.
:::

### 2024-07-05 paper.tex
Comparison with Shapley-Outlier result added.

\begin{Shaded}
\begin{Highlighting}[]
\FunctionTok{library}\NormalTok{(tourr)}
\FunctionTok{library}\NormalTok{(mulgar)}
\FunctionTok{set.seed}\NormalTok{(}\DecValTok{929}\NormalTok{)}
\NormalTok{vc\_null }\OtherTok{\textless{}{-}} \FunctionTok{matrix}\NormalTok{(}\FunctionTok{rep}\NormalTok{(}\FloatTok{0.5}\NormalTok{, }\DecValTok{5}\SpecialCharTok{*}\DecValTok{5}\NormalTok{), }\AttributeTok{ncol=}\DecValTok{5}\NormalTok{)}
\FunctionTok{diag}\NormalTok{(vc\_null) }\OtherTok{\textless{}{-}} \DecValTok{1}
\NormalTok{m\_null }\OtherTok{\textless{}{-}} \FunctionTok{rep}\NormalTok{(}\DecValTok{0}\NormalTok{, }\DecValTok{5}\NormalTok{)}
\NormalTok{vc\_samp }\OtherTok{\textless{}{-}} \FunctionTok{matrix}\NormalTok{(}\FunctionTok{rep}\NormalTok{(}\DecValTok{0}\NormalTok{, }\DecValTok{5}\SpecialCharTok{*}\DecValTok{5}\NormalTok{), }\AttributeTok{ncol=}\DecValTok{5}\NormalTok{)}
\FunctionTok{diag}\NormalTok{(vc\_samp) }\OtherTok{\textless{}{-}} \DecValTok{1}
\NormalTok{vc\_samp[}\DecValTok{4}\NormalTok{,}\DecValTok{5}\NormalTok{] }\OtherTok{\textless{}{-}} \SpecialCharTok{{-}}\FloatTok{0.47}
\NormalTok{vc\_samp[}\DecValTok{5}\NormalTok{,}\DecValTok{4}\NormalTok{] }\OtherTok{\textless{}{-}} \SpecialCharTok{{-}}\FloatTok{0.56}
\NormalTok{vc\_samp }\OtherTok{\textless{}{-}}\NormalTok{ vc\_samp}\SpecialCharTok{*}\FloatTok{0.1}
\NormalTok{m\_samp }\OtherTok{\textless{}{-}} \FunctionTok{c}\NormalTok{(}\DecValTok{0}\NormalTok{, }\DecValTok{0}\NormalTok{, }\DecValTok{0}\NormalTok{, }\FloatTok{1.9}\NormalTok{, }\FloatTok{2.3}\NormalTok{)}
\NormalTok{samp }\OtherTok{\textless{}{-}} \FunctionTok{as.data.frame}\NormalTok{(}\FunctionTok{rmvn}\NormalTok{(}\DecValTok{6}\NormalTok{, }
                           \AttributeTok{mn =}\NormalTok{ m\_samp, }
                           \AttributeTok{vc =}\NormalTok{ vc\_samp))}
\FunctionTok{animate\_xy}\NormalTok{(samp, }\FunctionTok{guided\_anomaly\_tour}\NormalTok{(}\FunctionTok{anomaly\_index}\NormalTok{(),}
  \AttributeTok{ellipse=}\NormalTok{vc\_null), }\AttributeTok{ellipse=}\NormalTok{vc\_null, }
  \AttributeTok{axes =} \StringTok{"bottomleft"}\NormalTok{, }\AttributeTok{half\_range=}\DecValTok{5}\NormalTok{, }\AttributeTok{center=}\ConstantTok{FALSE}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

@@

Here we will consider points to be outlying if they are more than
\(5 \sigma\) away from that mean value. With \(p=16\) this corresponds
to a Mahalanobis distance of \(60\) or larger. This will identify \(20\)
out of the \(68\) observations as outlying. Since these are outlying in
different combinations of variables, as found in \citet{filzmoser2018},
we will further separate the outlying points into clusters based on
similarity in direction. The similarity is computed by first normalizing
observations to have length \(1\) and then apply k-means clustering with
Euclidean distance. The cluster validation statistics computed using
\citet{fpc} suggest that \(k=4\) or \(k=5\) is reasonable, so for
simplicity we will work with \(k=4\).

The new anomaly index is first applied to the full dataset, such that
the final projection will be affected by averaging distance of points in
many directions. It provides a global picture showing where the outlying
points differ from the normal distribution.

@@

XXX SOMETHING ON WHAT THIS MEANS RELATIVE TO THE ORIGINAL EXAMPLE

>>>>

\begin{Shaded}
\begin{Highlighting}[]
\FunctionTok{library}\NormalTok{(tourr)}
\FunctionTok{animate\_xy}\NormalTok{(samp, }\FunctionTok{guided\_anomaly\_tour}\NormalTok{(}\FunctionTok{anomaly\_index}\NormalTok{(),}
  \AttributeTok{ellipse=}\NormalTok{vc\_null), }\AttributeTok{ellipse=}\NormalTok{vc\_null, }
  \AttributeTok{axes =} \StringTok{"bottomleft"}\NormalTok{, }\AttributeTok{half\_range=}\DecValTok{5}\NormalTok{, }\AttributeTok{center=}\ConstantTok{FALSE}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

@@

Here we will consider points to be outlying if they are more than
\(5 \sigma\) away from that mean value. With \(p=16\) this corresponds
to a Mahalanobis distance of \(60\) or larger. This will identify \(20\)
out of the \(68\) observations as outlying. Since these are outlying in
different combinations of variables, as found in \citet{filzmoser2018},
we will further separate the outlying points into clusters based on
similarity in direction. The similarity is computed by first normalizing
observations to have length \(1\) and then apply k-means clustering with
Euclidean distance. We use the Dunn index \citet{dunn}, computed using
\citet{fpc}, to select the preferred number of clusters, \(k=5\).

The new anomaly index is first applied to the full dataset, such that
the final projection will be affected by averaging distance of points in
many directions, see Fig. Figure~\ref{fig-weather} (a). It provides a
global picture showing where the outlying points differ from the normal
distribution. We can see that the contrast between maximum (atmx) and
minimum (atmn) temperature is primarily used in the direction where
points are most outlying. In the orthogonal direction the variables
precipitation (prc) and number of days with maximum temperature above
25°C (nsm) can be used to separate the clusters 2, 3 and 5. With this
solution the two points in cluster 4 cannot be resolved, they fall
inside the cluster. To better understand what is different for those two
years we separately run the anomaly index for this subset, see Fig.
Figure~\ref{fig-weather} (b). From the result we see that the maximum
temperature is not relevant for this cluster, but that they have low
values for the minimum temperature. We also see that there is still some
averaging, and the preferred solution is not placing either point on the
outside of the projected cluster.

@@

To compare our results to what is obtained by outlier detection methods,
we reproduce one of the results from \citet{filzmoser2018} in
Figure~\ref{fig-shapley}. On top we can see the cluster assignments for
all points that we have identified as outliers using the
\(p\)-dimensional Mahalanobis distance.

For example we see that cluster 4 contains the years 1962 and 1996, and
for these two years the result from \citet{filzmoser2018} also indicates
unusually low values for the minimum teperature, while there is
disagreement between outlyingness in other variables. Note that compared
to the years 1965 and 1966 they have unusually low minimum temperature
while having average values for the number of days with maximum
temparature above 25°C, and this contrast is also captured in the
projection.

When looking at the overall picture (Fig. Figure~\ref{fig-weather} (a)),
we can see that the linear projection is showing the difference between
minimum and maximum temperature, which is where clusters are outlying in
different directions. The direction of cluster 1 is where the average
maximum temperature is unusually higher compared to the minimum
temperature, while clusters 3 corresponds to years where the minimum
temperature is unusually close to the maximum temperature. This combined
information cannot be resolved with the cell detection algorithm, but we
see that for example the most recent years (cluster 3) were tagged for
having unusually high values for both the minimum and the maximum
temperature.

\includegraphics[width=6.25in,height=\textheight]{images/weather3.png}

\caption{\label{fig-shapley}Reproduction of the results from
\citet{filzmoser2018}, showing which cells have been identified as
outlying by their Shapley Cell Detector algorithm. Red cells indicate
values higher than expected, blue cells values lower than expected. For
years that have been identified as outliers in our approach we show the
cluster assignment at the top.}
